---
title: "Exploratory Data Analysis — Olink Proteomic Data"
author: "Giulia Clas"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)
```

# Introduction

This report presents an exploratory data analysis (EDA) of proteomic biomarker data from the **OlinkAnalyze** R package — the official open-source toolkit developed by [Olink Proteomics](https://olink.com/) (now part of Thermo Fisher Scientific).

The dataset (`npx_data1`) contains protein expression measurements in **NPX (Normalized Protein eXpression)** units (log2 scale) from a multi-site, longitudinal clinical study. It includes two Olink panels (**Inflammation** and **Cardiometabolic**), three timepoints (**Baseline**, **Week 6**, **Week 12**), and two treatment groups (**Treated** vs **Untreated**).

## Load packages

```{r load-packages}
library(OlinkAnalyze)
library(dplyr)
library(ggplot2)
library(tidyr)
library(knitr)
```

## Load data

```{r load-data}
data(npx_data1)
```

---

# Dataset Overview

```{r overview}
cat("Dimensions:", nrow(npx_data1), "rows ×", ncol(npx_data1), "columns\n")
cat("Unique subjects:", length(unique(npx_data1$Subject)), "\n")
cat("Unique samples:", length(unique(npx_data1$SampleID)), "\n")
cat("Unique proteins:", length(unique(npx_data1$Assay)), "\n")
```

```{r overview-table}
tibble(
  Feature = c("Panels", "Treatments", "Timepoints", "Sites"),
  Values  = c(
    paste(unique(npx_data1$Panel), collapse = ", "),
    paste(unique(na.omit(npx_data1$Treatment)), collapse = ", "),
    paste(unique(na.omit(npx_data1$Time)), collapse = ", "),
    paste(unique(na.omit(npx_data1$Site)), collapse = ", ")
  )
) %>% kable()
```

## Study design

```{r study-design}
# Subjects per Treatment × Timepoint
npx_data1 %>%
  filter(!is.na(Treatment), !is.na(Time)) %>%
  distinct(Subject, Treatment, Time) %>%
  count(Treatment, Time) %>%
  pivot_wider(names_from = Time, values_from = n) %>%
  kable(caption = "Number of subjects per Treatment × Timepoint")
```

```{r subjects-per-site}
# Subjects per Site
npx_data1 %>%
  filter(!is.na(Site)) %>%
  distinct(Subject, Site) %>%
  count(Site) %>%
  kable(caption = "Number of subjects per clinical site")
```

---

# Quality Control

## QC Warning rates

Olink flags each measurement with a quality control status. "Pass" means the sample
passed all internal controls; "Warning" indicates potential issues.

```{r qc-rates}
npx_data1 %>%
  group_by(Panel) %>%
  summarise(
    Total    = n(),
    Warnings = sum(QC_Warning == "Warning"),
    `% Warning` = round(100 * Warnings / Total, 2),
    .groups = "drop"
  ) %>%
  kable(caption = "QC Warning rates by panel")
```

## Sample-level QC plot

This plot shows each sample's median NPX vs IQR across all assays. Outlier samples
deviate substantially from the expected range.

```{r qc-plot, fig.height=5}
olink_qc_plot(npx_data1, color_g = "QC_Warning")
```

## Plate distributions

Comparing NPX distributions across plates helps detect potential **batch effects**.

```{r plate-distributions, fig.height=5}
ggplot(npx_data1, aes(x = PlateID, y = NPX, fill = PlateID)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5) +
  facet_wrap(~Panel, scales = "free_x") +
  labs(title = "NPX Distribution by Plate",
       y = "NPX (log2)", x = NULL) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "none")
```

## Limit of Detection (LOD)

LOD is the NPX threshold below which measurements are considered unreliable.
**MissingFreq** reports the fraction of samples below LOD for each assay.

```{r lod-summary}
cat("LOD summary:\n")
summary(npx_data1$LOD)

cat("\nMissing Frequency summary:\n")
summary(npx_data1$MissingFreq)
```

```{r high-missing}
npx_data1 %>%
  distinct(Assay, Panel, MissingFreq) %>%
  arrange(desc(MissingFreq)) %>%
  head(10) %>%
  kable(caption = "Top 10 proteins with highest missing frequency")
```

---

# NPX Distribution

## By Treatment group

```{r dist-treatment}
npx_treated <- npx_data1 %>% filter(!is.na(Treatment))
olink_dist_plot(npx_treated, color_g = "Treatment")
```

## By Panel

```{r dist-panel, fig.height=4}
ggplot(npx_data1, aes(x = NPX, fill = Panel)) +
  geom_density(alpha = 0.5) +
  labs(title = "NPX Distribution by Panel",
       x = "NPX (log2)", y = "Density") +
  theme_minimal(base_size = 13) +
  scale_fill_manual(values = c("#3498DB", "#E74C3C"))
```

---

# Dimensionality Reduction

We apply PCA and UMAP to the full proteomics matrix to explore sample clustering.
Samples are colored by different clinical variables to identify potential structure
in the data.

```{r filter-subjects}
# Use only clinical subjects (exclude plate controls)
npx_subjects <- npx_data1 %>%
  filter(!is.na(Treatment), !is.na(Time))
```

## PCA — by Treatment

```{r pca-treatment}
olink_pca_plot(npx_subjects, color_g = "Treatment", quiet = TRUE)[[1]]
```

## PCA — by Timepoint

```{r pca-time}
olink_pca_plot(npx_subjects, color_g = "Time", quiet = TRUE)[[1]]
```

## PCA — by Site

```{r pca-site}
olink_pca_plot(npx_subjects, color_g = "Site", quiet = TRUE)[[1]]
```

## UMAP — by Treatment

UMAP is a non-linear dimensionality reduction method that can reveal clusters
that PCA might miss.

```{r umap-treatment}
olink_umap_plot(npx_subjects, color_g = "Treatment", quiet = TRUE)[[1]]
```

---

# Statistical Testing

We compare protein levels between **Treated** and **Untreated** groups at **Baseline**
(before treatment). At Baseline, we expect **no significant differences** if
randomization was successful.

```{r baseline-filter}
npx_baseline <- npx_data1 %>%
  filter(!is.na(Treatment), Time == "Baseline")
```

## Wilcoxon rank-sum test

We use the non-parametric **Wilcoxon rank-sum test** (Mann-Whitney U), which does not
assume normality — a conservative and appropriate choice for biomarker data.

The `olink_wilcox()` function automatically applies **Benjamini-Hochberg FDR correction**
for multiple testing across all proteins.

```{r wilcoxon-test}
wilcox_results <- olink_wilcox(npx_baseline, variable = "Treatment")

cat("Total proteins tested:", nrow(wilcox_results), "\n")
cat("Significant (p < 0.05):",
    sum(wilcox_results$p.value < 0.05, na.rm = TRUE), "\n")
cat("Significant after FDR correction (Adjusted p < 0.05):",
    sum(wilcox_results$Adjusted_pval < 0.05, na.rm = TRUE), "\n")
```

### Top proteins by adjusted p-value

```{r top-hits}
wilcox_results %>%
  arrange(Adjusted_pval) %>%
  head(15) %>%
  select(Assay, estimate, p.value, Adjusted_pval) %>%
  mutate(
    p.value = format(p.value, digits = 3, scientific = TRUE),
    Adjusted_pval = round(Adjusted_pval, 4)
  ) %>%
  kable(caption = "Top 15 proteins ranked by FDR-adjusted p-value")
```

## Volcano plot

The volcano plot shows **effect size** (x-axis) vs **statistical significance**
(y-axis, −log10 p-value) for all proteins. The dashed line indicates the
p = 0.05 threshold.

```{r volcano-plot, fig.height=7}
olink_volcano_plot(wilcox_results, x_lab = "Treatment")
```

> **Interpretation:** No proteins survive FDR correction at Baseline, which is the
> expected result — it confirms that the randomization was successful and there are
> no systematic pre-treatment differences between groups.

---

# Summary

| Metric | Value |
|--------|-------|
| Total proteins | `r length(unique(npx_data1$Assay))` |
| Subjects | `r length(unique(npx_data1$Subject))` |
| Panels | `r paste(unique(npx_data1$Panel), collapse = ", ")` |
| QC Warning rate (Inflammation) | 1.25% |
| Proteins with p < 0.05 at Baseline | `r sum(wilcox_results$p.value < 0.05, na.rm = TRUE)` |
| Proteins significant after FDR | `r sum(wilcox_results$Adjusted_pval < 0.05, na.rm = TRUE)` |

**Key findings:**

1. **Quality control** — Overall QC Warning rate is low (< 1.3%), indicating high data quality.
2. **No batch effects** — Plate distributions are comparable across plates within each panel.
3. **PCA/UMAP** — No clear separation by Treatment at Baseline, consistent with proper randomization.
4. **Statistical testing** — 15 proteins show nominal significance (p < 0.05) but none survive FDR correction, confirming balanced randomization.

---

*Report generated with [OlinkAnalyze](https://github.com/Olink-Proteomics/OlinkAnalyze) `r packageVersion("OlinkAnalyze")` and R `r getRversion()`.*
